Principles for Open Data Curation: A Case Study with the New York City 311 Service Request Data

### 2024-06-21 manuscript/manuscript.tex
Review and editing through Part 10
Doing overall review, editing, and summarizing of text to reduce page size and # of figures/graphics. Making good progress for first sweep through. Have reviewed up to Section 10.

\subsection{Issues with Police Precincts} 	
\label{sec:police-precincts}
	A curious case also exists when examining the two nearly identical 
	fields - police\_precincts and police\_precinct. Both of those fields 
	are among the ``computed'' fields in the dataset. Using 
	\href{https://www.nyc.gov/site/nypd/bureaus/patrol/precincts-landing.page}
	{NYPD Precinct listings} it's possible to determine the valid 
	police precincts; they are generally numeric or have numeric 
	representations in the dataset. However, what we find is that 
	both fields police\_precinct and police\_precincts  have  35\% 
	invalid entries. Unfortunately, it's not the same invalid entries.

	For the police\_precincts field, there are 2,171,864 invalid entries 
	(35\%)., a consequential number of errors. Similarly for the 
	police\_precinct field, there are 2,171,778 invalid entries 
	(also 35\%).

@@

\subsection{created\_date and closed\_date(s) -- Negative Duration}
\label{sec:negativeduration}
	Let's look first at the created\_date and closed\_date fields. You might 
	expect that these two fields are automatically populated by the 
	SR application software. In such a case, creating an SR would 
	automatically trigger a created\_date based on a system clock. 
	Thus the created\_date is associated with setting an SR status 
	to ``new'' or ``open''. And similarly when an SR's 	status is 
	changed to ``closed'' then the closed\_date would be 
	populated. Unfortunately, that is not always the case. There are
	several anomalies with the various date fields.

@@

	This violin chart shows the broad spread of negative-duration SRs, albeit with 
	the extremely large negative-duration SR removed. While there are few 
	outlier points, the magnitude of the negative-durations is troubling and can 
	produce bizarre analytical results.

@@

\subsection{created\_date and closed\_date(s) --  Zero Duration}
\label{sec: zeroduration}		
	An even more serious, and more prevalent problem with closed\_date(s) 
	occurs when the closed\_date and created\_date are exactly the same -- to 
	the second -- which accordingly creates a \textbf{zero duration'}. Again 
	this is nonsensical, but the presence of such zero durations can again 
	severely distort analytics. There are 191,141 such SRs representing 3.1\% of 
	all non-blank data; not an insignificant number. As shown in the chart 
	below, 99\% of the zero duration SRs appear to predominately occur in five 
	Agencies (Dept of Mental Health \&Hygiene, Dept of Transportation, 
	Dept of Business, Dept of Sanitation, and Dept of Environmental Protection.) 
	This is not in-line with the overall Agency distribution of SRs and indicates an 
	Agency-specific issue, and the solution likely lies at those Agencies.

@@

\subsection{created\_date and closed\_date(s) -- Midnight \&Noon}
\label{sec:midnightandnoon}
	An additional problem discovered with the created\_date and 
	closed\_date fields; there appears to be an unusually large number of SRs 
	created or closed at exactly midnight (00:00:00) and exactly noon (12:00:00), 
	to the second. The distribution of SR creation and closure largely follows the 
	work-day clock with many SRs created during day-light hours, and 
	fewer SRs 	created at night and in the early hours of the morning. However, 
	here there appears to be significantly greater numbers of SR closed exactly 
	at midnight and exactly at noon, as well as a significant number of SRs 
	created exactly at noon. The reason behind this anomaly is worth investigating as 
	it affects the SR ``duration'' question. If SR durations are used to measure 
	responsiveness for the delivery of services, and if those times are not an 
	accurate reflection of the start/finish of the Service, then the use of duration 
	as a service quality metric is called into question. Observations during 
	this 2-year period include:

@@

	There is, of course, a broad distribution of the time that SRs are 
	created by day and hour, by night and day, by weekday and weekend.  
	One would think that that distribution would be (mostly) evenly distributed 
	by the minute and second. That is, if there are normally 100 SRs created 
	between 1-2pm, that the distribution to the minute and second would be 
	close to random. For example, there is no reason to suspect that a larger 
	number of SRs would be created at 6 minutes past the hour rather 
	than at 5 minutes past the hour. And certainly, if you look at the second 
	that these SRs are created you would again expect a (more-or-less) 
	random distribution; that there would be no reason to believe that a higher 
	(or lower) number of SRs would arrive at 18 seconds past the minute 
	versus at 39 seconds past the minute. Unfortunately, that turned out not to be true.

	Let's look at the SR creation distribution. We discovered that there 
	are a significant number of SRs created exactly at noon. To reveal this trend, 
	we looked at the exact date-time stamp of the created\_date(s). We began 
	by selected only SRs that were created exactly on the minute; at 09:00:00 
	or 13:00:00 for example.  Here is a visualization of the number of SRs 
	created and closed by hour-of-the-day:

	\begin{figure}[tbp]
		\centering
		\includegraphics[width = \textwidth]{Created_Hourly_SR_count.pdf}
		\caption{SRs by Hourly created\_date(s)}
		\label{fig:hourly-created}
	\end{figure}

	\begin{figure}[tbp]
		\centering
		\includegraphics[width = \textwidth]{Closed_Hourly_SR_count.pdf}
		\caption{SRs by Hourly closed\_date(s)}
		\label{fig:hourly-closed}
	\end{figure}

	The created\_date shows a maximum value at 12:00. The closed\_date chart 
	show significant spikes at both midnight and noon. The midnight (00:00) 
	spike appears to be an anomaly especially as compared to the hours just 
	before and after midnight where there is normally a pronounced slump 
	in SR closures. And both the closed-at-noon and closed-at-midnight are 
	approaching 2\textsigma's from the hourly average.

	Let's drill down on each of these outlier points. First, we aggregate the 
	created\_date over the 2-year period that fall exactly on the top-of-the-hour 
	with the minute and second values equal to 0, e.g. 09:00:00, 14:00:00. 
	This visualization reveals a clearer picture of where these spikes are:

	\begin{figure}[tbp]
		\centering
 		\includegraphics[width = \textwidth]
 		{2-year-trend-SR_created_by_top_of_hour.pdf}
		\caption{SRs Created at Top-of-the-Hour}
		\label{fig:tophourcreated}
	\end{figure}

	One other way to visualize this anomaly is to look at the busiest day 
	during this 2-year period (Friday, 2023-09-29). Here we will aggregated 
	the created\_date(s) by minute (with seconds equal to zero). This 
	provides a minute-by-minute look at SR creation on the busiest day 
	of this study. Here we see a clear spike at exactly noon (12:00), well 
	beyond the 3\textsigma line, and well above any other hour of the day.

@@

	Let's take a similar approach to the closed\_date distribution over the 
	2-year period. We aggregate SRs by close\_date to find the top-of-the-hour, 
	where minute and second values equal to 0, e.g. 20:00:00. Here we see 
	spikes at both midnight (00:00) and at noon (12:00), with the midnight 
	spike being well beyond the 3\textsigma line. Also note how much lower 
	(and nearly equal) all the other top-of-the-hour values are, 
	relative to the midnight and noon spikes.

	\begin{figure}[tbp]
		\centering
		\includegraphics[width = \textwidth]
		{2-year-trend-SR_closed_by_top_of_hour.pdf}
		\caption{SRs Closed at Top-of-the-Hour}
		\label{fig:tophourclosed}
	\end{figure}

	As before, let's visualize this anomaly by looking at the closed\_date(s) 
	on the busiest day during this 2-year study period (Friday, 2023-09-29). 
	We aggregate closed\_date(s) by the minute (with seconds equal to zero). 
	This gives a minute-by-minute view of SR closure on the busiest day 
	of this study. Clearly visual is the very large closure spikes occurring at 
	both midnight and noon. The chart highlights both of these anomalies 
	with spikes of 286 and 120 at 00:00:00 and 12:00:00 respectively. 
	Again, note how these spikes are so very much above the median 
	values for all other minutes.

@@

	These unusual patterns of created \& closed SRs at exactly the hours 
	of midnight and noon likely indicates the presence of a bulk create/close 
	software process; a process that perhaps automatically ``closes'' 
	(or ``creates'') a large number of SRs with a provided time-stamp of 
	midnight (00:00:00) or noon (12:00:00), and does so in bulk. 
	If so, then addressing this issue likely requires a closer look at those 
	Agencies providing these suspect SR closed/created times. The desirable 
	state is that the created\_date and closed\_date fields correctly 
	reflect the true opening and closing of the SRs, such that an 
	accurate ``duration'' can be computed for the SR life-cycle. 
	Otherwise, we are likely seeing durations that are artificially 
	manipulated by a software process performing bulk uploads 
	to the 311 system.

	Looking at the distribution by Agency for the ``closed-exactly-at-midnight'' 
	shows that \textgreater90\% of these suspect SRs come from just two 
	Agencies - Dept. of Buildings (DOB) and Dept. of Sanitation (DSNY). 
	If you include HPD and DOT, that accounts for 99\% of the suspect 
	SR closures. This Agency distribution does not follow the overall 
	SR distribution and we can assume that it is Agency specific.

@@

	However, the ``SR closed at noon'' visualization shows that  a single 
	Agency, DSNY, is responsible for \textgreater99\% of the 
	SRs ``closed-exactly-at-noon''. Accordingly, any effort to 
	resolve this issue necessitates working with DSNY to fully understand 
	how this is occurring.

@@

	The 99,779 SRs created-exactly-at-noon also appears to 
	originate with the Dept of Sanitation (DSNY) which is responsible 
	for \textgreater99\% of the suspect SRs.

	\begin{figure}[tbp]
		\centering
		\includegraphics[width = \textwidth]{created_at_noon_chart.pdf}
		\caption{SRs Created Exactly at Noon by Agency}
		\label{fig:noon-created}
	\end{figure}

	While it is difficult to ultimately know if these observations 
	are indeed anomalies; they could be the result of normal, 
	expected 311 SR activity. But at least by observation that 
	outcome appears very unlikely.

\subsection{resolution\_action\_update\_date}
\label{sec: resolutionaction}
	When an SR is updated, by any means on any field, the software 
	automatically populates the resolution\_action\_update\_date. 
	Analyzing that field it is apparent that some of the SR updates 
	are happening long, long after the SR is closed. In this dataset 
	there are eight SRs that have resolution\_action\_update\_date values 
	of 44,601 days after the closed\_date. The problem lies not with 
	the  resolution\_action\_update\_date, but rather with the closed\_date 
	values for these 8 SRs, all of which are in the format of ``1900-01-01''. 
	It has previously been discussed that these inaccurate dates may 
	possibly be coming from an Excel spreadsheet as that is the minimum 
	date which Excel can represent. Here is a sample of those SRs:

	\begin{table}[tbp]
	    \centering
	    \caption{Extremely long Resolution Updates}
		    \begin{tabular}{cccc}
		        \hline
		        Agency & Closed Date & Resolution Action Updated Date 
		        & Post Closed Update Duration (days) \\
		        \hline
		        DHS & 1900-01-01 & 2022-02-12 00:58:43 & 44602 \\
		        DHS & 1900-01-01 & 2022-02-11 14:28:59 & 44601 \\
		        DHS & 1900-01-01 & 2022-02-11 14:17:41 & 44601 \\
		        DHS & 1900-01-01 & 2022-02-11 13:45:59 & 44601 \\
		        DHS & 1900-01-01 & 2022-02-11 13:22:12 & 44601 \\
		        \hline
		    \end{tabular}
	    \label{tab:resolution-updates}
	\end{table}

	These obvious inaccurate date values are removed from the analysis. 
	But even so, there are some very late updates to SRs, with some 
	updates occurring almost two years after the SR is closed. There are a 
	total of 7460 SRs that are updated \textgreater30 days and \textless{}730 
	days after the closed\_date. The median of these late 
	(\textless{}30 days) post-closed resolution\_action\_update\_date(s) is 
	84 days. However, the mean resolution\_action\_update\_date for 
	all of the SRs is only 0.39 days (9.3 hours) after the SR is closed. 
	It is not known if this is normal behavior or an area that may 
	require further investigation.

	\begin{table}[tbp]
    	\centering
     \caption{Long post-closed Resolution Updates}
	    	\begin{tabular}{cccc}
		        \hline
		        Agency & Closed Date & Resolution Action Updated Date 
		        & Post Closed Update Duration (days) \\
		        \hline
		        TLC & 2023-05-25 14:21:12 & 2023-10-19 10:57:02 & 147 \\
		        TLC & 2023-07-12 11:03:46 & 2023-12-07 09:00:05 & 148 \\
		        TLC & 2023-08-03 10:34:23 & 2024-01-29 10:00:37 & 179 \\
		        TLC & 2022-04-04 10:54:39 & 2022-06-02 12:14:11 & 59 \\
		        TLC & 2022-12-05 15:23:45 & 2023-02-02 09:06:45 & 59 \\
		        \hline
		\end{tabular}
    	\label{tab:resolution}
	\end{table}

	Here is a look at the spread of the post-closed resolution\_action\_update\_date 
	values those SRs 	that are \textgreater30 days and \textless{}730 days. 
	There are some significant outliers:

@@

	If we look at a breakout of these (late?) update dates by Agency, 
	we can see that the vast majority li with two Agencies: Taxi \& Limo 
	Commission (TLC) and Dept of Sanitation (DSNY) which account 
	for 96\% of the SRs in question.  Is there a valid reason that TLC and 
	DSNY would be updating an SR over month (or months) after 
	the SR has been closed? Perhaps, but it seems to merit closer investigation.

	\begin{figure}[tbp]
		\centering
		\includegraphics[width = \textwidth]{postClosedBarChart.pdf}
		\caption{Post-Closed resolution\_action\_update\_date(s) 
		\textgreater30 days by Agency}
		\label{fig:resolution-by-agency}
	\end{figure}

\section{Accuracy and precision}
\label{sec:precision}
There is one area in particular where the question of precision vs. 
accuracy immediately arises, and that is with the Latitude and Longitude 
fields. Both the Longitude and Latitude fields are expresses as a 14-decimal 
number, e.g. 40.86769186022511 (also the Location field which is a 
straight concatenation of latitude and longitude). Given that 1 
degree of latitude at the equator is equal to 111.044736 kilometers, 
the ``1'' at the end of that number represents approximate 1.1104 nanometer 
or 1/1,000,000,000 of a meter. For reference a DNA molecule is 
approximately 2nm in width.

Clearly the representation of the Latitude and Longitude fields are a 
classic case of 14-digit precision, but limited accuracy. It is more likely 
that the Lat/Long values are accurate to the 5\textsuperscript{th} decimal place, 
about 3.64 feet; even that might be overstating things.

\section{Redundant \& Duplicate fields}\label{sec:duplicates}
During this analysis, several redundant fields were observed. Listed below 
is a discussion for each of a pair of data fields that are redundant or 
near-redundant. These fields should be examined further for 
possible consolidation in the overall dataset.

\subsection{latitude \& longitude and the location fields}
\label{sec:latlong}
The location field is a pure concatenation of the latitude 
and longitude fields, with a comma and parenthesis 
added. Example:

@@

The inclusion of the location field seems questionable as the data is 
arguably more difficult to extract than the two specific fields, especially 
for software programs.

@@

\subsection{borough and borough\_boundaries fields}
\label{sec:boroughboundaries}
The borough field is a text field for the five New York City boroughs (Manhattan, Queens, 
Brooklyn, etc.). The borough\_boundary field is numeric with values 
from 1-5 corresponding to the five boroughs (Staten Island, Brooklyn, 
Queens, Manhattan, Bronx). When the borough\_boundary field is 
translated and compared to the borough field, there is a 
98.3\% match. Given the size of the dataset, that still leaves 
110,715 non-matching values, not an insignificant number.

And that, in a nutshell, is the problem with two ``near-duplicate'' 
fields. It is like the  story of a man wearing two watches; he's never 
quite sure what time it is. And with ``near-duplicate'' fields, which 
field is correct when they disagree? If one field is blank, can you 
use the other one? Which field can be used as the reference value. And 
as we will see, there are several such examples in this dataset.

We can tell where the non-matches occur. It is mostly in DOT, NYPD, 
and DSNY which represent 90\% of the non-matching values.

\begin{figure}[tbp]
	\centering
	\includegraphics[width = \textwidth]{non_matching_borough_boundaries_chart.pdf}
	\caption{Non-matching borough and borough\_boundary by Agency}
	\label{fig:borough-boundaries}
\end{figure}

\subsection{borough and taxi\_company\_borough fields}
\label{sec:taxicompanyborough}
The two fields borough and taxi\_company\_borough might appear to be 
related. In fact, despite the names, the two data fields are nearly 
completely different with only 0.05\% matching. The mismatch is so 
high that it suggests the two fields are used very differently. This field 
is used exclusively by the Taxi \& Limo Commission which governs 
taxis and other cars for hire. There are a total of 3567 non-blank 
entries in the taxi\_company\_borough field, with 99.94\% blank.

 \subsection{incident\_zip and zip\_codes fields}
 \label{sec:zipcodes}
 The zip\_code field is one of the six computed fields and its validity 
 has been explored in Section \ref{sec:zip-codes} and found to be 
 lacking with 55.55\% of the entries invalid according to the 
 USPS, while the companion zipcode field incident\_zip field had 
 an accuracy rate of 99.93\%. It is not surprising that only 0.58\% of 
 the two fields match (37,292 matches out of 6.4 million SRs.). While 
 an Agency breakdown can be determined, given the questionable 
 computation of the zip\_code field, such an analysis would seem 
 to provide any useful information.

 \subsection{police\_precinct and police\_precincts fields} 
 \label{sec:police} 
Both the police\_precinct and police\_precincts are among the computed 
fields not cited in the Data Dictionary. Their usage however, is easy to 
discern. Are these two fields duplicates? Nearly so:  99.94\% of the entries 
match. Since the source of the fields and the underlying computational 
process are not know, it remains a challenge to determine which 
field is the ``right'' field to use. (The validity of these fields is 
covered in Section \ref{sec:police-precincts}.)

 \subsection{agency and agency\_name fields}
 \label{sec:agencyname}
 While not duplicates, these fields have a 1:1 correspondence.  The Agency 
 field contains abbreviations for the City agencies such as NYPD, DOT, 
 HPD, DEP, etc. The agency\_name field contains the full name of 
 the various organizations: New York Police Department, Department 
 of Transportation, Department of Housing Preservation and 
 Development, Department of Environmental Protection, and so on. Given 
 how well known the NYC Agency abbreviations are it seems redundant 
 to include both in the dataset.

\subsection{landmark and street\_name}
\label{sec:landmark}
The landmark field is listed in the Data Dictionary as ``Can refer to 
any noteworthy location, including but not limited to, parks, 
hospitals, airports, sports facilities, performance spaces, etc.'' We 
found that was not the case. To be sure many of the entries in 
the landmark field do contain landmark names, e.g.``Pennsylvania 
Station'', ``La Guardia Airport'', ``Gramercy Park'', the vast majority 
of the entries are street names, e.g. ``Fenton Avenue'', ``Steinway 
Street'', ``Broadway''. These entries appear to be very similar to 
those in the street\_name field. And in fact, the values of 
street\_name and landmark exactly match 62\% of the time, and 
while this is not a full duplicate, it is a precentage of matching 
that is certainly indicative of duplicate usage. Even the non-matches 
(excluding blanks) would appear to be matches except for minor 
spelling and nomenclature changes, e.g. ``NINTH AVE'' \& ``9 AVE''.

\begin{table}[ht]
    \centering
    \caption{Non-matches between 'street\_name' and 'landmark' fields}
	    \begin{tabular}{>{\raggedright\arraybackslash}p{5cm} >
	 	{\raggedright\arraybackslash}p{5cm}}
	 	\toprule
	      \textbf{street\_name} & \textbf{landmark} \\
	      \midrule
	        MACDOUGAL ST & MAC DOUGAL ST \\
	        NINTH AVE & 9 AVE \\
	        NORTH SIXTH ST & NORTH 6 ST \\
	        SOUTH FOURTH ST & SOUTH 4 ST \\
	        SIXTH AVE & 6 AVE \\
	        THIRD AVE & 3 AVE \\
	        EAST FIRST ST & EAST 1 ST \\
	        FOURTH AVE & 4 AVE \\
	        SAINT LAWRENCE AVE & ST LAWRENCE AVE \\
	        SOUTH FIRST ST & SOUTH 1 ST \\
	        WEST FIFTH ST & WEST 5 ST \\
	        MT HOPE PL & MOUNT HOPE PL \\
	        PENN STA & PENNSYLVANIA STA \\
	      \bottomrule
	    	\end{tabular}
    \label{landmark}
\end{table}

The distribution by Agency for the non-matches somewhat follows 
the overall SR distribution (NYPD, DPR, DSNY), but with some notable 
exceptions such as Taxi \& Limo Commission, Dept of Homeless 
Services (DHS), and Dept of Health \& Mental Hygiene (DOHMH).

\begin{figure}[tbp]
	\centering
  	\includegraphics[width = \textwidth]
  	{non-matchingstreet_nameandlandmark.pdf}
  	\caption{Non-matching street\_name \& landmark by Agency}
  	\label{fig:landmarkchart}
\end{figure}

@@

These two pairs of streets are used to help identify the location of the 
reported incident, as it is common in New York City to provide the 
street address and the cross street, as in ``24 W 90th street between 
Columbus and Amsterdam''. The Data Dictionary states that these 
streets are ``based on the geo validated incident location''. It then 
adds ``If the service request pertains to a location in the middle of a 
block, cross street will be provided by geocoding...If the service 
request pertains to a street corner, intersection will be provided by 
geocoding.''  One might think that the cross\_street(s) would apply 
for ``middle of a block'' incidents, while the intersection\_street(s) 
would apply to those incidents that happen on a street corner; 
that those two sets of fields would be exclusive to one another; 
an either/or situation.

Unfortunately, that is not the case. For the majority of data rows, both 
pair of fields are populated and are duplicates. In fact, 88\% of the 
time, cross\_street\_1 exactly matches intersection\_street\_1 and 
cross\_street\_2 exactly matches intersection\_street\_2, Only 
11.9\% are non-matches.

\textit{(Note: Address standardization was applied to the street 
addresses using the R \emph{campfin} package. This was done so as 
to prevent the software from generating a non-match between 
streets such as ``240 E 69\textsuperscript{th} ST'' and ``240 E 
69\textsuperscript{th} Street'' which are clearly the same address, 
just slight formatting and spelling differences.)}

The problem occurs when the two pairs of streets do not match, such as 
when one is blank and the other is not. Which street is correct? If one 
is blank, say cross\_street\_1, but  intersection\_street\_1 is not blank, should 
you use the non-blank one?  Let's look at some examples:

\begin{table}[tbp]
	\centering
     \caption{Matching/Non-Matching cross\_street\_1 \& intersection\_street\_1}
     	\label{tab:streets1}
	    	\begin{tabular}{l l l}
	      	\toprule
	      	\multicolumn{3}{c}
	      	{\textbf{Matching cross\_street\_1 \& intersection\_street\_1}} \\
	      	\midrule
	      	\textbf{cross\_street\_1} & \textbf{intersection\_street\_1} & \textbf{agency} \\
	      	\midrule
	       	FORT HAMILTON PKWY & FORT HAMILTON PKWY & NYPD \\
	        	87 ST              & 87 ST              & NYPD \\
	        	EAST 169 ST        & EAST 169 ST        & NYPD \\
	        	EAST 74 ST         & EAST 74 ST         & DOT  \\
	        	CHERRY AVE         & CHERRY AVE         & NYPD \\
	      	\midrule
	      	\multicolumn{3}{c}
	      	{\textbf{Non-matching cross\_street\_1 and intersection\_street\_1}} \\
	      	\midrule
	      	\textbf{cross\_street\_1} & \textbf{intersection\_street\_1} & \textbf{agency} \\
	      	\midrule
	        	KAPPOCK ST     & NETHERLAND AVE   & DOT \\
	        	GREENE AVE     & CLINTON AVE      & DOT \\
	        	LEWIS AVE      & FULTON ST        & DOT \\
	        	GERARD AVE     & EAST 161 ST      & DOT \\
	        	8 AVE          & WEST 136 ST      & DOT \\
	      	\bottomrule
	   	\end{tabular}
\end{table}

\begin{figure}[tbp]
	\centering
	\includegraphics[width = \textwidth]
	{non-matchingcross_street_1andintersection_street_1.pdf}
	\caption{Non-matching X\_Street\_1 \& Intersection\_Street\_1}
	\label{fig:xstreet1}
\end{figure}

>>>>

\subsection{Issues with Police Precincts} 	
\label{sec:police-precincts}
A curious case also exists when examining the two nearly identical 
fields - police\_precincts and police\_precinct. Both of those fields 
are among the ``computed'' fields in the dataset. Using 
\href{https://www.nyc.gov/site/nypd/bureaus/patrol/precincts-landing.page}
{NYPD Precinct listings} it's possible to determine the valid 
police precincts; they are generally numeric or have numeric 
representations in the dataset. However, what we find is that 
both fields police\_precinct and police\_precincts  have  35\% 
invalid entries. Unfortunately, it's not the same invalid entries. For 
the police\_precincts field, there are 2,171,864 invalid entries 
(35\%)., a consequential number of errors. Similarly for the 
police\_precinct field, there are 2,171,778 invalid entries 
(also 35\%).

@@

\subsection{created\_date and closed\_date(s) -- Negative Duration}
\label{sec:negativeduration}
For the created\_date and closed\_date fields, you might 
expect that these two fields are automatically populated by the 
SR application software. Thus the created\_date is associated with setting an SR status 
to ``new'' or ``open''. And similarly when an SR's 	status is 
changed to ``closed''  the closed\_date would be 
populated. Unfortunately, that does not appear to be the case as
there are several anomalies with the various date fields.

@@

This violin chart shows the broad spread of negative-duration SRs, albeit with 
the extremely large negative-duration SR removed. While there are few 
outliers, the magnitude of the negative-durations is troubling and can 
produce bizarre analytical results.

@@

\subsection{created\_date and closed\_date(s) --  Zero Duration}
\label{sec: zeroduration}		
A more serious and prevalent problem with closed\_date(s) 
occurs when the closed\_date and created\_date are exactly the same -- to 
the second -- which accordingly creates a \textbf{zero duration'}. Again 
this is nonsensical, and the presence of such zero durations can again 
severely distort analytics. There are 191,141 such SRs representing 3.1\% of 
all non-blank data. As shown in the chart 	below, 99\% of the 
zero duration SRs appear to predominately occur in five 
Agencies (Dept of Mental Health \&Hygiene, Dept of Transportation, 
Dept of Business, Dept of Sanitation, and Dept of Environmental Protection.) 
This is not in-line with the overall Agency distribution of SRs and indicates an 
Agency-specific issue, and the solution likely lies at those Agencies.

@@

\subsection{created\_date and closed\_date(s) -- Midnight \&Noon}
\label{sec:midnightandnoon}
Another problem discovered with the created\_date and 
closed\_date fields - there appears to be an unusually large number of SRs 
created or closed at exactly midnight (00:00:00) and exactly noon (12:00:00), 
to the second. The distribution of SR creation and closure largely follows the 
work-day clock with many SRs created during day-light hours, and 
fewer SRs 	created at night and the early hours of the morning. However, 
here there appears to be significantly greater numbers of SR closed exactly 
at midnight and exactly at noon, as well as a significant number of SRs 
created exactly at noon. Observations during this 2-year period include:

@@

One other way to visualize this anomaly is to look at the busiest day 
during this 2-year period (Friday, 2023-09-29). Then aggregate 
the created\_date(s) by minute (with seconds equal to zero), providing
a minute-by-minute look at SR creation. Here we see a clear 
spike at exactly noon (12:00), well beyond the 3\textsigma line.

@@

As before, let's visualize this anomaly by looking at the closed\_date(s) 
on the busiest day during this 2-year study period (Friday, 2023-09-29). 
We aggregate closed\_date(s) by the minute (with seconds equal to zero). 
The chart highlights anomalies with spikes at 00:00:00 and 
12:00:00 respectively.

@@

These unusual patterns of created \& closed SRs at exactly the hours 
of midnight and noon likely indicates the presence of a bulk create/close 
software process that perhaps automatically ``closes'' 
(or ``creates'') a large number of SRs with a provided time-stamp of 
midnight (00:00:00) or noon (12:00:00). Again, such behavior will
distort the duration of these SRs. Looking at the distribution 
by Agency for the ``closed-exactly-at-midnight'' 
shows that \textgreater90\% of these suspect SRs come from just two 
Agencies - Dept. of Buildings (DOB) and Dept. of Sanitation (DSNY).

@@

However, the ``SR closed at noon'' visualization shows that  a single 
Agency, DSNY, is responsible for \textgreater99\% of the 
SRs ``closed-exactly-at-noon''.

@@

\subsection{resolution\_action\_update\_date}
\label{sec: resolutionaction}
When an SR is updated, by any means on any field, the 311 software 
automatically populates the resolution\_action\_update\_date. 
Analyzing that field it becomes apparent that some of the SR updates 
are happening long after the SR is closed. There are a 
total of 7460 SRs that are updated \textgreater30 days and \textless{}730 
days after the closed\_date (to avoid infeasible dates such as 1900-01-01).
It is not known if this is normal behavior or an area that may 
require further investigation.

@@

\section{Accuracy and precision}
\label{sec:precision}
An area where the question of precision vs. accuracy arises is
with the Latitude and Longitude fields. Both the Longitude and 
Latitude fields are expresses as a 14-decimal 
number, e.g. 40.86769186022511. Also the Location field which is a 
straight concatenation of latitude and longitude). Given that 1 
degree of latitude at the equator is equal to 111.044736 kilometers, 
the ``1'' at the end of that number represents approximate 1.1104 nanometer 
or 1/1,000,000,000 of a meter. For reference a DNA molecule is 
approximately 2nm in width. Clearly the representation of 
the Latitude and Longitude fields are a classic case of 14-digit 
precision, but limited accuracy.

\section{Redundant \& Duplicate fields}\label{sec:duplicates}
During this analysis, several redundant fields were observed. These 
fields should be examined further for possible consolidation.

\subsection{latitude \& longitude and the location fields}
\label{sec:latlong}
The location field is a pure concatenation of the latitude 
and longitude fields, with a comma and parenthesis 
added. The inclusion of the location field seems questionable 
as the data is arguably more difficult to extract than the 
two specific fields, especially for software programs. Example:

@@

@@

\subsection{borough and borough\_boundaries fields}
\label{sec:boroughboundaries}
The borough field is a text field for the five New York City boroughs (Manhattan, Queens, 
Brooklyn, etc.). The borough\_boundary field is numeric with values 
from 1-5 corresponding to the five boroughs (Staten Island, Brooklyn, 
Queens, Manhattan, Bronx). When the borough\_boundary field is 
translated and compared to the borough field, there is a 
98.3\% match. And that is the problem with two ``near-duplicate'' 
fields; which field is correct when they disagree?

\subsection{borough and taxi\_company\_borough fields}
\label{sec:taxicompanyborough}
The two fields borough and taxi\_company\_borough might appear to be 
related. In fact, despite the names, the two data fields are nearly 
completely different with only 0.05\% matching. The mismatch is so 
high that it suggests the two fields are used very differently. This field 
is used exclusively by the Taxi \& Limo Commission which governs 
taxis and other cars for hire.

 \subsection{incident\_zip and zip\_codes fields}
 \label{sec:zipcodes}
 The zip\_code field is one of the six computed fields and its validity 
 has been explored in Section \ref{sec:zip-codes} and found to be 
 lacking with 55.55\% of the entries invalid according to the 
 USPS, while the companion zipcode field incident\_zip field had 
 an accuracy rate of 99.93\%.

 \subsection{police\_precinct and police\_precincts fields} 
 \label{sec:police} 
Both the police\_precinct and police\_precincts are among the computed 
fields not cited in the Data Dictionary. Their usage however, is easy to 
discern. These two fields are near-duplicates with 99.94\% of the entries 
matching. Since the source of the fields and the underlying computational 
process are not know, it remains a challenge to determine which 
field is the ``right'' field to use. (The validity of these fields is 
covered in Section \ref{sec:police-precincts}.)

 \subsection{agency and agency\_name fields}
 \label{sec:agencyname}
 While not duplicates, these fields have a 1:1 correspondence.  The Agency 
 field contains abbreviations for the City agencies such as NYPD, DOT, 
 HPD, etc. The agency\_name field contains the full name of 
 the various organizations: New York Police Department, Department 
 of Transportation, Department of Housing Preservation and 
 Development, and so on. It seems redundant to include both.

\subsection{landmark and street\_name}
\label{sec:landmark}
The landmark field is listed in the Data Dictionary as ``Can refer to 
any noteworthy location, including but not limited to, parks, 
hospitals, airports, sports facilities, performance spaces, etc.'' Our analysis
found that was not the case. To be sure many of the entries in 
the landmark field do contain landmark names, e.g.``Pennsylvania 
Station'', ``La Guardia Airport'', etc., the vast majority 
of the entries are street names, e.g. ``Fenton Avenue'', ``Steinway 
Street'', ``Broadway''. These entries are similar to those in the 
street\_name field. And in fact, street\_name and landmark 
match 62\% of the time. While a high enough percentage of matching 
to be indicative of duplicate usage. Even the non-matches 
(excluding blanks) would appear to be matches except for minor 
spelling and nomenclature changes, e.g. ``NINTH AVE'' \& ``9 AVE''.

\begin{table}[ht]
    \centering
    \caption{Non-matches between 'street\_name' and 'landmark' fields}
	    \begin{tabular}{>{\raggedright\arraybackslash}p{5cm} >
	 	{\raggedright\arraybackslash}p{5cm}}
	 	\toprule
	      \textbf{street\_name} & \textbf{landmark} \\
	      \midrule
	        MACDOUGAL ST & MAC DOUGAL ST \\
	        NINTH AVE & 9 AVE \\
	        SIXTH AVE & 6 AVE \\
	        EAST FIRST ST & EAST 1 ST \\
	        FOURTH AVE & 4 AVE \\
	        SAINT LAWRENCE AVE & ST LAWRENCE AVE \\
	        MT HOPE PL & MOUNT HOPE PL \\
	        PENN STA & PENNSYLVANIA STA \\
	      \bottomrule
	    	\end{tabular}
    \label{landmark}
\end{table}

@@

These two pairs of streets are used to help identify the location of the 
reported incident, as it is common in New York City to provide the 
street address and the cross street, as in ``24 W 90th street between 
Columbus and Amsterdam''. For the majority of data rows, both 
pair of fields are populated and are duplicates; 88\% of 
cross\_street\_1 matches intersection\_street\_1. Similarly with
cross\_street\_2 exactly matches intersection\_street\_2.

\textit{(Note: Address standardization was applied to the street 
addresses using the R \emph{campfin} package. This was done 
to prevent generating a non-match between 
streets such as ``240 E 69\textsuperscript{th} ST'' and ``240 E 
69\textsuperscript{th} Street'' which are clearly the same address
but with formatting and spelling differences.)}

In addition to the matching/non-matching values, we encountered a number 
of \emph{near-matches}, cases where the 
cross\_street and intersection\_street are nearly identical, 
e.g. HARMON DR and HARMON RD. To identify these 
near-matches we measured the \href{https://en.wikipedia.org/wiki/Hamming_distance}
{Hamming Distance} between the two fields. The Hamming Distance is a 
measure of how many letters have to be changed in order for the two 
fields to match. For this analysis, we chose a Hamming Distance of 2. Here 
are the results for both cross\_street and intersection\_street.

\begin{table}[tbp]
    \centering
     \caption{Near-matches for cross\_street\_1, intersection\_street\_1}
     \label{tab:x1nearmatches}
		\begin{tabular}{l l l r}
	        \toprule
	        \textbf{cross\_street\_1} & \textbf{intersection\_street\_1} 
	        & \textbf{agency} & \textbf{hamming\_distance} \\
	        \midrule
	        WEST 168 ST    & WEST 167 ST           & DOT    & 1 \\
	        105 AVE        & 107 AVE               & DOT    & 1 \\
	        145 ST         & 150 ST                & DOT    & 2 \\
	        138 ST         & 139 ST                & DOT    & 1 \\
	        19 AVE         & 20 AVE                & DOT    & 2 \\
	        76 ST          & 79 ST                 & DOT    & 1 \\
	        67 ST          & 68 ST                 & DOT    & 1 \\
	        \bottomrule
	    \end{tabular}
\end{table}
